```{r}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with dplyr function using the star wars dataset:

```{r}
starwars
head(starwars)
?starwars
```

# Subset rows with `filter`

[**`filter()`**]{.underline} allows us to subset the rows of a dataset based on the values.

```{r}
## Filter for rows only for Star Wars Human Characters
filter(starwars, species == "Human")

## save to a new T object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

## Filter multiple columns at ounce
filter(starwars, species =="Human", sex == "male")
```

We can use different comparison operators to filter:

-   `!` is "not"

-   `==` is "equal"

-   `&` is "and"

-   `<` is "less than"

-   `>` is "greater than"

-   `is.na()` returns TRUE if value is NA

```{r}
## return Characters with a Missing Mass
filter(starwars, is.na(mass))
```

```{r}
## How to fer all Characters that are NOT Human?
filter(starwars, species != "Human")
```

**`variable_name %in% vector_of_values`** returns TRUE if the value off variable_name is equal to one of the values in vector_of_values

This is a shortcut for variable_name == vector_of_values[1] \| variable_name == vector_of_values[2]...

```{r}
skin_color <- "grey"
skin_color %in% c("yellow","fair","tan") #returns FALSE because "grey" is not in vector of colors
# "yellow" %in% c("yellow","fair","tan")

## How would we filter for characters with Brown or Blond Hair?
filter(starwars, hair_color %in% c("blond", "brown"))
```

**`grepl("characters_of_interest", variable_name)`** returns TRUE if the value of varibale-name contains the characters_of_interest

```{r}
filter(starwars, grepl("blond", hair_color))

## How would we get all Characters with Blue Eyes?
filter(starwars, grepl("blue", eye_color))
```

You can use `<`, `>`, `<=`, `>=` to compare Numeric Values or Categorical Variables

-   Nominal Categorical Variables (character vectors) are ranked alphabetically

-   Ordinal Categorical Variable (factors) are ranked based on the built-in-order

```{r}
## StarWars Characters with a Name starting with letters before "D"
filter(starwars, name < "D") #ranking is based on Alphabetically Order

str(diamonds)
filter(diamonds, cut > "Good") #ranking is based on Levels of the Ordered factor Cut

"D" < "d"
"d" < "D" #true
```

**`between(x, left, right`**) is a shorthand for x \> = left & x \<= right

```{r}
between(5, 1, 7) # TRUE because 5 is less than or eual to 7 and greater than or equal to 1
filter(starwars, between(mass, 50, 100))
```

### Practice Part 1

**Find all Star Wars Characters that**

1.  ***Are missing a hair color***

```{r}
starwars
filter(starwars, hair_color == "none")
filter(starwars, is.na(hair_color))
```

2.  ***Are shorter than 100cm and are not Droids***

```{r}
starwars
filter(starwars, height < 100, species != "Droid")
```

3.  ***Were born between the Years 100 and 200***

```{r}
starwars
filter(starwars, between(birth_year, 100, 200))
```

4.  ***Another of your own Choice (If Time)***

```{r}
#Characters Names that does not have a "A" in their
starwars
filter(starwars, !grepl("A", name))


#Characters that has at least 1 Star Ship
filter(starwars, length(starships) > 0)

#Charcters that are only Droids
filter(starwars, species =="Droid")
```

## Order Rows with `arrange`

**`arrange()`** reorders Rows based on the values of column(s)

```{r}
## Order by Character Names
arrange(starwars, name)
"a" < "B"
# Order of the Characters --->  aAbBcCdD...

## Order by Species and Birth Year
arrange(starwars, species, birth_year)

## use desc() to get the Birth Years in a Decending Order
arrange(starwars, species, desc(birth_year))
```

### Practice Part 2

1.  ***Arrange the Star Wars Data Set from Tallest to Shortest***

```{r}
starwars
arrange(starwars, desc(height))
```

2.  ***Arrange the Star Wars Data Set from Shortest to Tallest by Species***

```{r}
arrange(starwars, species, height) # Order by species, whithin  species order shrtest to tallest
arrange(starwars, height, species) # Order by Height, Tiebreaker is Alphabetic Species
```

3.  ***Alphabetize the Star Wars characters by their hair color then eye color***

```{r}
arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can select multiple columns in a row

```{r}
starwars
# All Columns beerween name and eye_color (inclusive)
select(starwars, name:eye_color)
```

We can also remove columns by name using **`-`**

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

#### Helpful Functions

**`everything()`** is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # Put Species first than the others

## Alphabetize Columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

**`stars_with("abc")`** matches column names that starts with **`("abc")`**

**`ends_with("abc")`** matches column names that end with **`("abc")`**

**`contains("abc")`** matches column names that contains **`("abc")`**

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

### Practice Part 3

1.  Select each Character's homeworld and species. What question might this subset of variables help answer?

-   #s of Species in a Planet/ types of Species in a Planet/ Etc

```{r}
starwars
select(starwars, name, homeworld, species)
```

2.  Brainstorm as many ways as possible to select `hair_color`, `skin_colo`r, `eye_color`, and `birth_year`

```{r}
starwars
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, hair_color:birth_year)
select(starwars, c(hair_color, skin_color, eye_color, birth_year))
select(starwars, c(hair_color:birth_year))
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
select(starwars, -c(name:mass, sex:starships))
```

## Add new Variables with **`mutate`**

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create Multiple Variables at ounce and we can refer to Variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## Only Keep the NEW Variables
transmute(starwars_small,
          height_m = height/100,
          bmi = mass/height_m^2)
```

**Summary Statistics** gives one Value for the entire column: **`sum()`**, **`mean()`**

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group by a Variable using the **`.by`** parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm =T),
       .by = species)
```

Change the type of a Variable

```{r}
mutate(starwars_small,
       mass = as.character(mass))

mutate(starwars_small,
       mass = as.logical(mass))
```

### Practice Part 4

1.  Convert Mass to Pounds [***(1kg = 2.2 lbs)***]{.underline}

```{r}
starwars
mutate(starwars_small,
       pound_lbs = mass*2.2)
```

## Summarize Data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

## Group by Species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create

**`n()`** is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

## Group by multiple Groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species, sex))
```

## Renaming

rename(data, new_name, = old_name)

```{r}
starwars
new_starwars <- rename(starwars, 
       birthYear = birth_year)

new_starwars
```
